Air Pollution from Electricity Production

### 2025-05-23 Paper_Endogenous.tex

\author{Lauren Beatty\thanks{Environmental Defense Fund  \hspace{.5cm} E-mail: lbeatty1@edf.org \hspace{.5cm}Website: \href{https://lbeatty1.github.io}{https://lbeatty1.github.io}}\\
Matthias Fripp\thanks{Environmental Defense Fund}}

@@

\newpage
\section{Introduction}
Climate change mitigation will require deep cuts to greenhouse gas emissions from the electricity sector.  Least-cost electricity planning models are powerful tools that help policymakers and capacity planners find portfolios of generation technologies that meet reliability requirements and emissions goals.

However, these models have historically ignored some of the external costs of electricity generation.  
 One major external costs is the generation of local air pollutants.  There is a large body of work on the health effects of air pollution, and a large body of work on the effects of electricity generation on air pollution. 
Air pollution is linked with numerous health impacts such as chronic obstructive pulmonary disease (COPD), acute lower respiratory illness, cerebrovascular disease, ischaemic heart disease, and lung cancer. \cite{Lelieveld2015TheScale} estimate that power generation is linked with 31$\%$ of premature mortality caused by outdoor air pollution in the U.S.

There is also work that has shown that air pollution exposures vary drastically across racial and socioeconomic groups.   \cite{Thind2019FineGeography} find that Black and non-Hispanic white people are exposed to more pollution than other groups and that exposures for low-income people are higher than for high-income people, even after controlling for race.  This highlights the importance for developing planning methods that alleviate rather than reinforce disproportionate burdens.

@@

Least-cost electricity planning models are the gold-standard models used to model capacity expansion in policy.  However, these models tend to have a narrow focus on minimizing costs -- potentially at the expense of distributional concerns.  The energy transition will have wide-ranging effects --such as shifts in the spatial distribution of jobs, or changes in air pollution, that are of interest to policy-makers, but are currently unincorporated into least-cost models.

@@

This paper makes an important contribution to the environmental justice literature by ...

\section{Data}
\subsection{Emissions Data}
Emissions data comes from two sources: EPA's Clean Air Markets Program Data (CAMD) and EPA's National Emissions Inventory (NEI).  The CAMD was created to track compliance with clean air programs, and thus, tracks emissions of CO$_2$, NO$_x$, and SO$_2$.  The NEI was created to track criteria pollutants, criteria precursors, and hazardous air pollutants from \textit{all} sources.  I get emissions of NH$_3$, VOC, and PM2.5 from the NEI.

@@

\section{Results}

\begin{figure}
\centering
\subcaptionbox{Mean empirical emissions rates by technology in tons per MWh.}{\includegraphics[width=0.45\textwidth]{Figures/EndogenousResults/base_short/base_short_emissions_by_technology.png}}%
\hfill
\subcaptionbox{Mean coefficients describing how a marginal MWh of generation affects mean group-level PM2.5 exposure.}{\includegraphics[width=0.45\textwidth]{Figures/EndogenousResults/base_short/base_short_exposure_by_technology.png}}%
\hfill
\caption{Unsurprisingly, coal is far dirtier than other technologies.  However, the discrepancy between technologies is more narrow for exposure.}
\label{EmissionsExposureCoefs}
\end{figure}
In figure \ref{EmissionsExposureCoefs} we plot both average emissions rates, and average exposure coefficients for various technologies.

@@

>>>>

\author{Lauren Beatty\thanks{Environmental Defense Fund  \hspace{.5cm} E-mail: lbeatty1@edf.org \hspace{.5cm}Website: \href{https://lbeatty1.github.io}{https://lbeatty1.github.io}}\\
Matthias Fripp\thanks{Energy Innovations}}

@@

\newpage
\section{Introduction}
Climate change mitigation requires deep cuts to greenhouse gas emissions from the electricity sector.  Least-cost electricity planning models are powerful tools that help policymakers and capacity planners find portfolios of generation technologies that meet reliability requirements and emissions goals.

However, these models have historically ignored some of the external costs of electricity generation. One major external costs is the generation of local air pollutants.  There is a large body of work on the health effects of air pollution, as well as a large body of work on the effects of electricity generation on air pollution. 
Air pollution is linked with numerous health impacts such as chronic obstructive pulmonary disease (COPD), acute lower respiratory illness, cerebrovascular disease, ischaemic heart disease, and lung cancer. \cite{Lelieveld2015TheScale} estimate that power generation is linked with 31$\%$ of premature mortality caused by outdoor air pollution in the U.S.

There is also work that has shown that air pollution exposures vary drastically across racial and socioeconomic groups.   \cite{Thind2019FineGeography} find that Black and non-Hispanic white people are exposed to more pollution than other groups and that exposures for low-income people are higher than for high-income people, even after controlling for race.  This highlights the importance of developing planning methods that alleviate rather than reinforce disproportionate burdens.

@@

Least-cost electricity planning models are the gold standard models used to model capacity expansion in policy.  However, these models tend to have a narrow focus on minimizing costs -- potentially at the expense of distributional concerns.  The energy transition will have wide-ranging effects --such as shifts in the spatial distribution of jobs, or changes in air pollution, that are of interest to policy-makers, but are currently unincorporated into least-cost models.

@@

This paper makes an important contribution to the environmental justice literature by endogenizing air pollution within a least-cost planning model, and proposing a framework to equitably reduce air pollution. Our findings indicate that major improvements to air pollution can be made with relatively small increases in overall cost.

\section{Data}
The fist task in our method is to construct emissions rates for sources in the model in tons per MWh. These emissions rates are constructed by combining emissions data from the U.S. Environmental Protection Agency (EPA), and generations data from the U.S. Energy Information Administration (EIA). We use the EPA-EIA Power Sector Data Crosswalk to merge generators across the two sources \citep{HuettemanEPA-EIACrosswalk}.

\subsection{Emissions Data}
Emissions data comes from two sources: EPA's Clean Air Markets Program Data (CAMD) and EPA's National Emissions Inventory (NEI).  The CAMD was created to track compliance with clean air programs, and thus, tracks emissions of CO$_2$, NO$_x$, and SO$_2$.  The NEI was created to track criteria pollutants, criteria precursors, and hazardous air pollutants from \textit{all} sources.  I get emissions of NH$_3$, VOC, and PM2.5 from the NEI.  Air pollution data is used to calculate source emissions rates in tons per MWh.

\subsection{Generation Data}
Data on the generation and location of existing plants comes from the EIA.

@@

\section{Results}
\subsection{Model Inputs}
There are a few findings from the construction of model inputs that are worth discussing. First, unsurprisingly, empirical data on emissions and generation reveals that coal is a dirty source of generation. However, when we translate emissions into marginal exposure coefficients, we find that the average relative difference between coal and natural gas decreases. In particular, it appears that natural gas plants are on average, closer to Asian and Latino population centers. A unit of electricity generated from coal only causes slightly more damage (in terms of population-weighted PM2.5 exposure) to Asians and Latinos than a unit of electricity generated from natural gas. In figure \ref{EmissionsExposureCoefs} we plot the average empirical emissions rates by technology in panel (a), and the mean coefficients describing how one MWh of generation affects mean group-level exposure of PM2.5 in $\mu$g/m$^3$.

This start divergence between the relative emissions rates and the relative exposure coefficients highlights one of the central premises of the paper -- that where generation is placed can have large impacts on pollution exposure, and that selectively choosing what generation goes where can lead to large decreases in pollution exposure at low costs.

\begin{figure}
\centering
\subcaptionbox{Mean empirical emissions rates by technology in tons per MWh.}{\includegraphics[width=0.45\textwidth]{Figures/EndogenousResults/base_short/base_short_emissions_by_technology.png}}%
\hfill
\subcaptionbox{Mean coefficients describing how a marginal MWh of generation affects mean group-level PM2.5 exposure.}{\includegraphics[width=0.45\textwidth]{Figures/EndogenousResults/base_short/base_short_exposure_by_technology.png}}%
\hfill
\caption{Unsurprisingly, coal is far dirtier than other technologies.  However, the discrepancy between technologies is more narrow for exposure.}
\label{EmissionsExposureCoefs}
\end{figure}

\subsection{Model Results}
Our high-level, key finding is that even a modest budget of $\$50$ billion through 2050 could pay for massive reductions in equitable air quality improvement. Our results have features that you would expect based on economic theory. There are increasing costs of reducing air pollution -- the gains from moving from least-cost to allocating $\$50$ billion are much larger than the gains from moving from allocating $\$50$ billion to $\$100$ billion. Our results also suggest that at the optimum, the min-max objective generates equal average exposures for all groups except Native Americans, who have consistently lower exposure levels.

@@

\subsection{Comparison with carbon-only cuts}
While we think these results have important policy implications, namely that improving air quality equitably can happen at low costs, it's important to note that our results don't point to which policies would achieve the optimal solution. Rather, these results are prescriptive -- \textit{"Build X, Y, and Z in regions A,B, and C to achieve the deepest equitable cuts in air pollution at a certain cost."}

Previous papers have found that carbon policies tend to decrease both total air pollution exposure, and the inequalities in air pollution exposure. This begs the question, how does a portfolio that minimizes air pollution exposure compare to a least-cost portfolio aimed at reducing carbon?

### 2025-05-23 Slides_3_6.tex

\usepackage[utf8]{inputenc}
\usepackage{csquotes}
\usepackage{latexsym,amsmath,xcolor,multicol,booktabs,calligra, animate, subfig}
\usepackage{graphicx,pstricks,listings,stackengine, bbm}
\usepackage{tikz}
\usetikzlibrary{fit,tikzmark}   
\usepackage{booktabs,cellspace}
\usepackage{color, colortbl}
\usepackage{hyperref}
\usepackage{color}
%Information to be included in the title page:
\title{Designing Equitable Electricity Systems}
\author{Lauren Beatty\\ \href{mailto:lbeatty@edf.org}{lbeatty@edf.org}\\
\vspace{.5cm}\\
Matthias Fripp\\
\href{mailto:mfripp@edf.org}{mfripp@edf.org}}
\institute{Environmental Defense Fund}
\date{May 6, 2024}

@@

\logo{\includegraphics[height=1cm]{EDF_color_webRGB_2022 (2).png}\hspace*{.03\paperwidth}\vspace{.01\paperwidth}}

@@

\begin{frame}
\frametitle{Introduction}
\textbf{Main Question:  How do we design electrical systems that center equity and justice \textit{without} sacrificing efficiency?}
\vspace{1cm}

\paragraph{\textbf{Previous Paradigm}}
\begin{itemize}
    \item Potentially harmful and polluting infrastructure is sited disproportionately near poor and marginalized communities.
    \item Planning decisions are informed by \textit{least cost} models.
    \item Communities are left in the lurch by technological changes.
\end{itemize}
\end{frame}

@@

{\nologo
\begin{frame}{Previous Attempts to Shoehorn Justice Into these Models}
\begin{center}
\begin{tikzpicture}[node distance=2cm]

\node(obj)[process, below of=start, text width=8cm]{\shortstack[l]{$\min \sum \text{Costs} + \text{Monetized Air Pollution Impacts}$ \\ \textit{ 
   s.t.   }\text{Everyone gets electricity and nothing breaks }}};

\end{tikzpicture}
\end{center}

\begin{itemize}
    \item Monetized impacts don't consider unjust distributional consequences.
    \item Equity weights can be difficult to generate and defend.
\end{itemize}

\end{frame}}

{\nologo
\begin{frame}{Our Approach}
    \begin{center}
\begin{tikzpicture}[node distance=2cm]

@@

\begin{frame}{Our Approach}
    \begin{enumerate}
        \item Use EPA emissions data to calculate emissions rates (Pounds per MWh)
        \item Use InMap Source-Receptor Matrix to calculate how one MWh of generation for each resource affects population-weighted average exposure for different groups.
        \item Use Switch to calculate least cost planning costs.
        \item Modify and run Switch to minimize air pollution exposure subject to a cost constraint.
    \end{enumerate}

@@

    The effect of deviations from the least-cost optimum will depend on the shape of the objective function.  This is an \textit{empirical} question!\\
    \vspace{.5cm}
    $\Rightarrow$ Perhaps you can get large improvements in health or jobs at only modest increases in cost.
\end{frame}

@@

    Please feel free to contact me at:
    \href{mailto:lbeatty@edf.org}{lbeatty@edf.org}\\
    \vspace{1cm}
\end{frame}

\begin{frame}{Bonus Slide!}
    This project started as a side-project to the Model Intercomparison Project -- a project aimed at harmonizing the inputs and comparing multiple open-source electricity planning models.

    \vspace{.4cm}

    \includegraphics[width=0.3\textwidth]{Figures/Output/ISRM_deathrate_by_scenario_2030.jpg}
    \includegraphics[width=0.3\textwidth]{Figures/Output/ISRM_deathrate_by_scenario_2040.jpg}
    \includegraphics[width=0.3\textwidth]{Figures/Output/ISRM_deathrate_by_scenario_2050.jpg}

\begin{itemize}
    \item Death rates vary widely across racial groups.
    \item Aggressive carbon policies reduce pollution exposure the fastest.
    \item Optimistic -- even under current policies future generations should have much better air quality.
\end{itemize}

\end{document}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{csquotes}
\usepackage{latexsym,amsmath,xcolor,multicol,booktabs,calligra, animate, subfig}
\usepackage{graphicx,pstricks,listings,stackengine, bbm}
\usepackage{tikz}
\usetikzlibrary{fit,tikzmark}   
\usepackage{booktabs,cellspace}
\usepackage{color, colortbl}
\usepackage{hyperref}
\usepackage{color}
%Information to be included in the title page:
\title{Air Pollution from Electricity Production}
\author{Lauren Beatty\\ \href{mailto:lbeatty@edf.org}{lbeatty@edf.org}}
\institute{Environmental Defense Fund}
\date{\today}

@@

\logo{\includegraphics[height=.5cm]{EDF_color_webRGB_2022 (2).png}\hspace*{.03\paperwidth}\vspace{.01\paperwidth}}

@@

\begin{frame}
\frametitle{Introduction}
Two Different Questions / Approaches:
\begin{enumerate}
    \item Downscaling
    \begin{itemize}
        \item \textbf{Question}: How might a particular plan for decarbonization affect air pollution.
        \item \textbf{Method}: Sequentially run a least-cost planning model, then take the outputs of that model and predict pollution.
    \end{itemize}
    \item Endogenous
    \begin{itemize}
        \item \textbf{Question}: What decarbonization plan most dramatically decreases air pollution equitably?
        \item \textbf{Method}: Alter the planning model to minimize pollution subject to a cost constraint.
    \end{itemize}
\end{enumerate}
\vspace{1cm}

@@

\begin{frame}{Background on Least-Cost Electricity Planning Models: Pros and Cons}
    \begin{itemize}
        \item Powerful for out-of-sample questions

        \item Persuasive in policy arguments.

        \item No or very little backwards validation -- We have no evidence that these models make good predictions about what will happen.

        \item Rely on strong assumptions about future costs and demand, as well as the strong assumption that these are known ex-ante.
    \end{itemize}
\end{frame}

\begin{frame}{Air Pollution Modelling}
\textbf{Translate emissions at points to concentrations over space.}

\vspace{.3cm}

Multiple models of varying complexity. The main models used in policy work are:
\begin{itemize}
    \item APEEP
    \item INMAP
    \item EASIUR
\end{itemize}

\vspace{.3cm}

These are all \textbf{average-annual models}, with reduced-form representations of more complex chemical transport models. Each answers the question: \textit{How much does pollution affect average annual exposure?} Therefore, they are suitable for thinking about policies affecting aggregate emissions, but not suitable for thinking about things like hourly NAAQs.
\end{frame}

\begin{frame}{Downscaling}
\begin{enumerate}
    \item Run least-cost planning model
    \item Model outputs annual generation for each ``resource," such as Combined Cycle Plants in New England ISO North.
    \begin{itemize}
        \item How much resources are aggregated is a user choice which needs to balance computational feasibility and model accuracy.
    \end{itemize}
    \item Translate model outputs into point source locations and aggregate emissions of important pollutants (in InMAP -- PM2.5, NOx, SOx, VOCs, NH3).
        \begin{itemize}
            \item Involves lots of assumptions about how to allocate generation to existing plants, and the location of new plants.
        \end{itemize}
    \item Run those aggregate emissions through an air pollution model of your choice.
    \item Translate concentrations into health impacts.
\end{enumerate}
\end{frame}

\begin{frame}{MIP Results}
    \begin{columns} % Create two columns
        \column{0.5\textwidth} % First column: 50% of slide width
        \centering
        \includegraphics[width=\linewidth]{Figures/Output/full-base-200/GenX/ISRM_2027_TotalPM25_concentrationmap.jpg}
        \captionof{figure}{PM2.5 Concentrations in 2027 in the base-200 scenario}

        \column{0.5\textwidth} % Second column: 50% of slide width
        \centering
        \includegraphics[width=\linewidth]{Figures/Output/full-base-200/GenX/ISRM_2050_TotalPM25_concentrationmap.jpg}
        \captionof{figure}{PM2.5 Concentrations in 2050 in the base-200 scenario}
    \end{columns}
\end{frame}

\begin{frame}{MIP Results}
    \includegraphics[width=\textwidth]{Figures/Output/Compare_scenarios_current-base_GenX.jpg}
\end{frame}

{\nologo
\begin{frame}{Endogenous Exposure}
    \begin{center}
\begin{tikzpicture}[node distance=2cm]

@@

\begin{frame}{Our Approach}
    \begin{enumerate}
        \item Use EPA emissions data to calculate emissions rates (Pounds per MWh)
        \item Use InMap Source-Receptor Matrix to calculate how one MWh of generation for each resource affects population-weighted average exposure for different groups.
        \item Use Switch to calculate least-cost planning costs.
        \item Modify and run Switch to minimize air pollution exposure subject to a cost constraint.
    \end{enumerate}

@@

    The effect of deviations from the least-cost optimum will depend on the shape of the objective function.  This is an \textit{empirical} question!\\
    \vspace{.5cm}
    $\Rightarrow$ Perhaps you can get large improvements in health at only modest increases in cost.
\end{frame}

\begin{frame}{Flavor of Results}
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/EndogenousResults/current_policies_short_simplified/exposure_cost_PPF.png}
\end{frame}

@@

    Please feel free to contact me at:
    \href{mailto:lbeatty@edf.org}{lbeatty@edf.org}\\
    \vspace{1cm}
\end{frame}
\end{document}
